---
title: "Andrew's Bacteria-Phage Correlations"
author: "Rachel Rodgers"
date: "4/8/2021"
output: 
  html_document:
    code_folding: hide
editor_options: 
  chunk_output_type: console
  
---

```{r load-libraries, message=FALSE, warning=FALSE}
library("pheatmap")
library("psych")
library("phyloseq")
library("tidyverse")
```

Function definitions.

```{r fxn-dfs}
FillEmptyRanks <- function(physeqObj) {
  # Function replaces missing values (NA) or values of "not classified"
  #   in a phyloseq's tax_table with a more informative name, generally
  #   "unclassified <previousTaxonomicRankName" or "unclassified" if no
  #   previous rank is classified.
  #
  # Args:
  #   physeqObj: An object of class Phyloseq
  # Returns:
  #   Phyloseq object with modified taxonomy table

  # Extract taxonomy table from physeq object, convert any factor columns to character
  taxTableOld <- as.data.frame(as(tax_table(physeqObj), "matrix"))
  taxTableModified <- mutate(taxTableOld, across(where(is.factor), as.character))
  
  for (i in 1:nrow(taxTableModified)) {
    
    lastString <- "unclassified"
    
    for (j in 1:ncol(taxTableModified)) {
      # value is not NA or "not classified", set lastString variable
      if (!is.na(taxTableModified[i, j]) & taxTableModified[i, j] != "not classified") {
        lastString <- taxTableModified[i, j]
        # value is NA and is in column 1
        } else if (is.na(taxTableModified[i, j]) & j == 1) {
          taxTableModified[i, j] <- lastString
        # value is NA and is in column 2+
        } else if (is.na(taxTableModified[i, j]) & j > 1) {
          taxTableModified[i, j] <- paste("unclassified", lastString)
        # value is equal to "not classified"
        } else if (taxTableModified[i, j] == "not classified") {
          if (lastString == "not classified") {
            taxTableModified[i, j] <- "unclassified"
            } else {
              taxTableModified[i, j] <- paste("unclassified", lastString) 
            }
        }
    }
  }

  # Convert the new taxonomy table back into class tax_table, overwrite tax slot,
  #   and assign to new phyloseq object
  newTaxonomy <- as(taxTableModified, "matrix")
  physeqModifiedTaxonomy <- physeqObj
  
  tax_table(physeqModifiedTaxonomy) <- tax_table(newTaxonomy)
  return(physeqModifiedTaxonomy)
  
}

PivotData <- function(df, rank) {
  # Pivots melted phyloseq data frame into wide format
  #
  # Args:
  #   df: Melted phyloseq data
  #   rank: The taxonimc rank of interest
  #
  # Returns:
  #   wide-formatted data frame from melted phyloseq data
  dfModified <- df %>%
    select(Sample, Abundance, !!sym(rank)) %>%
    pivot_wider(names_from = all_of(rank), values_from = "Abundance") %>%
    mutate(across(where(is.numeric), ~ replace(., is.na(.), 0)))
}

MakeCorrelationMatrix <- function(abdTable, bacteria, phage, sig_p = 0.05) {
  # Generate a correlation matrix, significant correlation matrix and matrix
  #   of correlation p-values from an abundance data frame. Returns a list obj.
  #
  # Args:
  #   abdTable - data frame of abundance information (sample names are row names,
  #   column names are taxa)
  #   bacteria - vector of all bacteria names in current abd table
  #   phage - vector of phage names in current abd table
  #   sig_p - significance value for correlations
  # Returns:
  #   correlationList$correlation_matrix - correlation matrix containing all 
  #       correlations regardless of significance
  #   correlationList$sig_correlation_matric - correlation matrix containing
  #       only the significant correlations
  #   correlationList$correlationPValues - matrix of p-values for all correlations
  emptyCols <- colnames(abdTable)[colSums(abdTable) == 0]
  keepBacteria <- bacteria[!bacteria %in% emptyCols]
  keepPhage <- phage[!phage %in% emptyCols]
  sigCorrMtx <- NA

  #correlation <- cor(x = abdTable, method = "spearman")
  correlation <- corr.test(x = abdTable, method = "spearman", adjust = "fdr")
  
  corrReduced <- correlation$r %>%
    as.data.frame() %>% 
    rownames_to_column(var = "row_taxa") %>%
    filter(row_taxa %in% keepBacteria) %>% # only bacteria will be in the rows
    dplyr::select(row_taxa, all_of(keepPhage)) # only phage will be in the columns
  
  corrMtx <- column_to_rownames(corrReduced, "row_taxa") 
    
  # correlation p-values (filtered)
  corrPValReduced <- correlation$p %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "row_taxa") %>%
    filter(row_taxa %in% keepBacteria) %>% # only bacteria will be in the rows
    dplyr::select(row_taxa, all_of(keepPhage)) # only phage will be in the columns
    
  corrPValMtx <- column_to_rownames(corrPValReduced, var = "row_taxa")
  # pivot the correlations and p-values into a long data frame, use to generate
  #   significant correlation matrix
  corrPivot <- corrReduced %>% 
    pivot_longer(cols = !row_taxa, names_to = "column_taxa", 
                 values_to = "correlation") %>% 
    unite(col = "comparison", row_taxa, column_taxa, sep = "-")
  
  pValPivotLUT <- corrPValReduced %>% 
    pivot_longer(cols = !row_taxa, names_to = "column_taxa",
                 values_to = "p_value") %>% 
    unite(col = "comparison", row_taxa, column_taxa, sep = "-") %>% 
    deframe()
  
  # Filter to just keep correlations that were significant, remaining non-sig
  #   comparisons will fill with NA
  sigCorrPivot <- corrPivot %>% 
    mutate("p_value" = pValPivotLUT[comparison]) %>% 
    filter(p_value <= sig_p)
  
  if (dim(sigCorrPivot)[1] != 0) { # if not empty
    sigCorrMtx <- sigCorrPivot %>% 
      separate(col = "comparison", into = c("row_taxa", "column_taxa"), sep = "-") %>% 
      select(-p_value) %>% 
      pivot_wider(names_from = column_taxa, values_from = correlation) %>% 
      column_to_rownames(var = "row_taxa")
  }
  
  #corrReduced <- correlation %>%
    #as.data.frame() %>% 
    #rownames_to_column(var = "row_taxa") %>%
    #filter(row_taxa %in% keepBacteria) %>% # only bacteria will be in the rows
    #dplyr::select(row_taxa, all_of(keepPhage)) # only phage will be in the columns
  
  correlationList <- list("correlation_matrix" = corrMtx,
                          "sig_corraltion_matrix" = sigCorrMtx,
                          "correlation_p_values" = corrPValMtx)
  
}

PlotCorrelationHeatmap <- function(correlationMtx,  title = NULL) {
  # Calls pheatmap on given correlationMtx.  Does not cluster rows or columns,
  #  sets empty columns to grey color.
  #
  # Args:
  #   correlationMtx - a correlation matrix
  #   title - plot title
  # Returns:
  #   pheatmap graphic
  correlationHeatmap <- pheatmap(correlationMtx, na_col = "grey", main = title,
                                 cluster_rows = FALSE, cluster_cols = FALSE)
}
```

Loading bacterial and phage phyloseq objects.  Changing taxonomic categories for phage from lower to upper-case for ease.

```{r load-data, results='hide'}
physeqBacteria <- readRDS("psm.DS3.RDS")
sampleDataBacteria <- as(sample_data(physeqBacteria), "data.frame")
physeqBacteria 

physeqPhage <- readRDS("psm.DS3.phage.RDS") # currently contains phage + other stuff
sampleDataPhage <- as(sample_data(physeqPhage), "data.frame")
physeqPhage

# Adjust the capitalization of the phage taxonomy table for ease (currently lower case)
phageRanksOld <- colnames(tax_table(physeqPhage))
phageRanksNew <- str_to_title(phageRanksOld)
colnames(tax_table(physeqPhage)) <- phageRanksNew
```

Determine which samples are seroconversion Yes and No so heatmaps can be divided appropriately.

```{r sample-seroconversion-status}
# Since the bacteria and phage data have different numbers of samples, we need
#   to make sure we identify all the samples in each seroconversion category

bacteriaSero <- sampleDataBacteria %>% 
  rownames_to_column(var = "sample_name") %>% 
  select(sample_name, seroconversion) # 95 samples
bacteriaSeroYes <- bacteriaSero %>% 
  filter(seroconversion == "Yes") %>% 
  pull(sample_name)
bacteriaSeroNo <- bacteriaSero %>% 
  filter(seroconversion == "No") %>% 
  pull(sample_name)

phageSero <- sampleDataPhage %>% 
  rownames_to_column(var = "sample_name") %>% 
  select(sample_name, seroconversion) # 84 samples
phageSeroYes <- phageSero %>% 
  filter(seroconversion == "Yes") %>% 
  pull(sample_name)
phageSeroNo <- phageSero %>% 
  filter(seroconversion == "No") %>% 
  pull(sample_name)

# merge the samples from phage and bacteria
seroYesSamples <- unique(c(bacteriaSeroYes, phageSeroYes))
seroNoSamples <- unique(c(bacteriaSeroNo, phageSeroNo))
```

Phyloseq objects will be glommed at all taxonomic ranks.  Checking the number of missing taxonomic ranks (NA's) in these columns so uninformative names in the taxonomy table can be replaced.

```{r checking-for-missing-taxonomic-names}
ranks <- c("Order", "Family", "Genus")
names(ranks) <- ranks

taxTableBacteria <- as.data.frame(as(tax_table(physeqBacteria), "matrix"))
numMissingBacterialRanks <- map_int(taxTableBacteria, ~ sum(is.na(.)))
numMissingBacterialRanks[ranks] 

taxTablePhage <- as.data.frame(as(tax_table(physeqPhage), "matrix"))
numMissingPhageRanks <- map_int(taxTablePhage, ~ sum(is.na(.)))
numMissingPhageRanks[ranks] 
```

We need to fill in these missing taxonomy names before proceeding so correlation matrices are more informative.  After correcting missing taxonomy, we can proceed to modifying the data to generate matrices of abundances for correlation tests.

```{r get-bacterial-data}
# Adjust taxonomy
physeqBacteriaModified <- FillEmptyRanks(physeqBacteria) 

# Melt physeq at various ranks, and remove taxa having < 0.01 abundance
bacteriaAbundanceDFList <- map(.x = ranks,
                               .f = ~ physeqBacteriaModified %>%
                                 tax_glom(.x) %>%
                                 transform_sample_counts(function(x){x/sum(x)}) %>%
                                 psmelt() %>% filter(Abundance > 0.01))

# Pivot the abundance tables to make the columns taxa, the rows samples, and the
#   values abundances
bacteriaMeltList <- vector(mode = "list", length = length(bacteriaAbundanceDFList))

for (i in 1:length(bacteriaAbundanceDFList)) {
  
  currentDF <- bacteriaAbundanceDFList[[i]]
  currentRank <- names(bacteriaAbundanceDFList)[i]
  
  meltedDF <- PivotData(df = currentDF, rank = currentRank)
  
  bacteriaMeltList[[i]] <- meltedDF
  names(bacteriaMeltList)[i] <- currentRank
}
```

```{r get-phage-data}
# Adjust taxonomy
physeqPhageModified <- FillEmptyRanks(physeqPhage) 

# Melt, remove taxa that are not in the Phage superkingdom from physeqPhageModified
phageAbundanceDFList <-  map(.x = ranks,
                             .f = ~ physeqPhageModified %>% 
                               tax_glom(.x) %>% 
                               transform_sample_counts(function(x){x/sum(x)}) %>% 
                               subset_taxa(Superkingdom == "Phage") %>% 
                               psmelt() %>% 
                               filter(Abundance > 0.01))

# Pivot
phageMeltList <- vector(mode = "list", length = length(phageAbundanceDFList))

for (i in 1:length(phageAbundanceDFList)) {
  
  currentDF <- phageAbundanceDFList[[i]]
  currentRank <- names(phageAbundanceDFList)[i]
  
  meltedDF <- PivotData(df = currentDF, rank = currentRank)
  
  phageMeltList[[i]] <- meltedDF
  names(phageMeltList)[i] <- currentRank
}
```

Merge melted abundance tables from phage and bacteria together by Sample column.  Also store the names of the phage and bacteria at each taxonomic level so we can organize the heatmaps to avoid redundant rows/columns.

```{r merge-abundance-tables}
# Merge abundance tables, and store all unique phage and bacteria names for
#   each merged rank
mergedAbundanceTables <- vector(mode = "list")
phageNames <- vector(mode = "list") # for organizing heatmaps
bacteriaNames <- vector(mode = "list") # for organizing heatmaps

for (i in 1:length(bacteriaMeltList)) {
  
  # current bacteria data
  bacteriaData <- bacteriaMeltList[[i]]
  bacteriaRank <- names(bacteriaMeltList)[i]
  bacteriaTaxa <- names(bacteriaData)[names(bacteriaData) != "Sample"]
    
  for (j in 1:length(phageMeltList)) {
    
    # get index
    idx <- length(mergedAbundanceTables)
    
    # current phage data
    phageData <- phageMeltList[[j]]
    phageRank <- names(phageMeltList)[j]
    phageTaxa <- names(phageData)[names(phageData) != "Sample"]
    
    # merge the data, store
    mergedName <- paste0("b_", bacteriaRank, "-p_", phageRank)
    mergedData <- merge(bacteriaData, phageData, by = "Sample", all = FALSE) %>%
      column_to_rownames(var = "Sample")
    
    # update index value and store data
    idx <- idx + 1
    
    mergedAbundanceTables[[idx]] <- mergedData
    names(mergedAbundanceTables)[idx] <- mergedName
    
    bacteriaNames[[idx]] <- bacteriaTaxa
    names(bacteriaNames)[idx] <- mergedName
    
    phageNames[[idx]] <- phageTaxa
    names(phageNames)[idx] <- mergedName
    
    print(idx)
    print(mergedName)
    
  }
}
```

Run correlations:

```{r correlationMatrices, message=FALSE, warning=FALSE}
correlationMatrices <- vector(mode = "list", 
                              length = length(mergedAbundanceTables))

for (i in 1:length(mergedAbundanceTables)) {

  currentComparison <- names(mergedAbundanceTables)[i]
  currentAbdTable <- mergedAbundanceTables[[i]]
  
  # Get the bacteria and phage names for the current comparison
  allBacteria <- bacteriaNames[[currentComparison]]
  allPhage <- phageNames[[currentComparison]]
  
  # Split currentAbdTable into seroYes and seroNo
  currentAbdTable <- currentAbdTable %>%
    rownames_to_column(var = "sample_name")
  
  seroYesAbdTable <- currentAbdTable %>% 
    filter(sample_name %in% seroYesSamples) %>% 
    column_to_rownames(var = "sample_name")
  
  seroNoAbdTable <- currentAbdTable %>% 
    filter(sample_name %in% seroNoSamples) %>% 
    column_to_rownames(var = "sample_name")

  abdTableList <- list("seroYes" = seroYesAbdTable, "seroNo" = seroNoAbdTable)
  
  correlationMtxList <- map(abdTableList, MakeCorrelationMatrix,
                            bacteria = allBacteria,
                            phage = allPhage)
  
  # Store correlation matrices in list
  names(correlationMatrices)[i] <- currentComparison
  correlationMatrices[[i]] <- correlationMtxList
}
```

```{r save-corrMtx-and-correlationPValues, eval=FALSE}
outPath <- "./output_DS3"
dir.create(outPath)

for (i in 1:length(correlationMatrices)) {

  comparison <- names(correlationMatrices)[i]
  correlationMtxList <- correlationMatrices[[i]]
  
  # correlation matrices
  titlesCorrMtx <- as.list(paste(comparison, names(correlationMtxList),
                                 "Correlation_Matrix.txt", sep = "_"))
  
  correlations <- list(correlationMtxList$seroYes$correlation_matrix,
                       correlationMtxList$seroNo$correlation_matrix)
  names(correlations) <- titlesCorrMtx
  correlationSavePaths <- paste0(outPath, names(correlations))
  walk2(.x = correlations, .y = correlationSavePaths,
        .f = ~ write.table(x = .x, file = .y, quote = FALSE, sep = "\t",
                           row.names = TRUE, col.names = TRUE))
  
  # p-value matrices
  titlesPVal <- as.list(paste(comparison, names(correlationMtxList),
                              "Correlation_P-Values.txt", sep = "_"))
  pValues <- list(correlationMtxList$seroYes$correlation_p_values,
                  correlationMtxList$seroNo$correlation_p_values)
  names(pValues) <- titlesPVal
  pvalueSavePaths <- paste0(outPath, names(pValues))
  walk2(.x = pValues, .y = pvalueSavePaths,
        .f = ~ write.table(x = .x, file = .y, quote = FALSE, sep = "\t",
                           row.names = TRUE, col.names = TRUE))
  
  
}

```

## Plots of all correlations: 

```{r plot-all-heatmaps}
# Plot all heatmaps
for (i in 1:length(correlationMatrices)) {
  currentComparison <- names(correlationMatrices)[i]
  correlationMtxList <- correlationMatrices[[i]]
  
  #print(currentComparison)
  #print(names(correlationMtxList))
  
  titles <- as.list(paste(currentComparison, names(correlationMtxList)))
  names(titles) <- names(correlationMtxList)
  
  for (j in 1:length(correlationMtxList)) {
    PlotCorrelationHeatmap(correlationMtx = correlationMtxList[[j]]$correlation_matrix,
                           title = titles[j])
  }
}
```

## Plots of only significant correlations:

```{r plot-only-sig-heatmaps}
# Plot only significant correlation heatmaps

for (i in 1:length(correlationMatrices)) {

  currentComparison <- names(correlationMatrices)[i]
  correlationMtxList <- list("seroYes" = correlationMatrices[[i]]$seroYes$sig_corraltion_matrix,
                             "seroNo" = correlationMatrices[[i]]$seroNo$sig_corraltion_matrix)
  
  titles <- as.list(paste(currentComparison, names(correlationMtxList)))
  names(titles) <- names(correlationMtxList)
  
  for (j in 1:length(correlationMtxList)) {

    isEmpty <- is.na(correlationMtxList[[j]])
    
    if (isEmpty == TRUE) {
      message(paste("No significant correlations for ", titles[j]))
    } else {
      numRows <- nrow(correlationMtxList[[j]])
      numCols <- ncol(correlationMtxList[[j]])
      tableSize <- numRows*numCols
      
      if (tableSize > 1) {
        
        PlotCorrelationHeatmap(correlationMtx = correlationMtxList[[j]],
                               title = titles[j])
        
        
      } else {
        message(paste("Only one significant correlation value for ", titles[j],
                      "cannot plot with pheatmap."))
      }
    }
  }
}

    
```
